Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-02-05 SCP2008/paper/metamonads.tex
partial changes, on the way to reorganize the paper

@@

The monad represents fine-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code. The ML module system
turns out to be most convenient.

@@

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
We will be using MetaOCaml operations to build code expressions, values
of type |('a,'v) code|. However, we wish to make our generators
abstract over particular code representations. Therefore, when writing
the LU generator (Section~\ref{sec:det} and below), instead of
MetaOCaml brackets and escapes we will be using the code combinators
described below. These combinators are functions that produce and
combine values of the type of \emph{abstract code}.

@@

Various aspects of our LU generator such as determinant, rank, 
and permutation tracking may need to keep a state during
code generation. For example, the initializing section of the LU
generator will invoke a function (|decl|) of the determinant aspect 
which ``declares'' whether tracking the determinant is needed, 
so that the
latter may generate let-binding for reference cells that track the
magnitude and the sign of the determinant. The generator of the
row-swapping code will notify the determinant aspect so that the
latter may insert code that flips the sign of the tracked
determinant. The final section of the LU generator will ask the
determinant aspect to generate code producing the final, signed value
of the determinant. Clearly the determinant aspect needs to maintain
state from one invocation of its functions to another; this state should
include the ``names'' of mutable variables accumulating the sign and the
magnitude of the determinant.

The simplest method of keeping the state of the generator is by using
mutable variables, private to the Determinant module (aspect). That,
however, makes our aspects stateful. Although we are generating
imperative code, we would like to keep our generators stateless and
purely functional, for ease of comprehension and reasoning.  Our main
program may include several LU generators referring to one Determinant
aspect -- which may be present in one shared instance or in
several. That is of no concern if the Determinant module is
stateless. With stateful modules, the issues of aliasing or separate
instantiation are the source of very subtle problems.

We therefore chose a different way of maintaining the generator state,
using a monad. We already need a monad for let-insertion; we add state as
an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
actions |fetch| and |store| to access that monadic state. The state is
threaded throughout the entire code-generation computation.

This monadic state has to accommodate several distinct pieces of state, for
various aspects. We should be able to add a new aspect -- which may need
to keep its own state as part of the overall monadic state -- without
modifying or even recompiling the rest of the code. Thus our monadic
state should be extensible. We could use an extensible
record: an OCaml object. Each aspect would have its own field; record
subtyping would assure modularity. Alas, this approach makes it
difficult to create an initial state, to pass to the monad's |runM|
method. We would be required to know the names of all fields and should know
the proper initial value for these fields, which breaks modularity.

let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generation-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.

\subsection{Determinant aspect}
\label{sec:det}
To track the determinant, we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign (|upd_sign|) or the absolute
value of the determinant, and converting the tracking state
to the final determinant value of type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| (see definition below) where |outdet| is |unit|
and all the
functions of that module generate no code. For integer matrices, and
in general whenever the matrix elements do not form a field, we
have to track some aspects of the determinant, even if we do not output
it, as the fraction-free update needs those aspects. The determinant tracking
aspect is complex because tracking variables, if any, are to be declared at the
beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. 
Other aspects follow a similar outline.

>>>>

Various aspects of our generator will need to keep a state during
code generation (see section~\ref{sec:det} for an example).

The simplest method of keeping the state of the generator is by using
mutable variables, private to each module (aspect). That would however make our
aspects stateful. Although we are generating imperative code, we would like to
keep our generators stateless and purely functional, for ease of comprehension
and reasoning.  Our main
program may include several generators referring to one particular
aspect -- which may be present in one shared instance or in
several. That is of no concern if the module is
stateless, but with stateful modules, the issues of aliasing or separate
instantiation are a source of very subtle problems.

We therefore chose a different way of maintaining generator state,
using a monad. We already need a monad for let-insertion; we now add state as
an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
actions |fetch| and |store| to access that monadic state. The state is
threaded throughout the entire code-generation computation.

This monadic state has to accommodate several distinct pieces of state, for
various aspects. We should be able to add a new aspect -- which may need
to keep its own state as part of the overall monadic state -- without
modifying or even recompiling the rest of the code. Thus our monadic
state should be extensible. We could use an extensible
record: an OCaml object. Each aspect would have its own field; record
subtyping would insure modularity. Alas, this approach makes it
difficult to create an initial state, to pass to the monad's |runM|
method. We would be required to know the names of all fields and should know
the proper initial value for these fields, which breaks modularity.

let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generation-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.  Our approach has the advantage of generating much
clearer error messages.

@@

The monad represents fine-scale code generation. We need tools for
larger-scale modularization; conveniently, we can use whatever abstraction
mechanisms we want to structure our code generators, as long as these
abstractions do not infiltrate the generated code. For our purposes, the
ML module system turns out to be most convenient.

\jacques{next paragraph will need rewritten when this section has
finally taken shape}
In this section, we will go through various aspects of one algorithm
and explain how they are implemented.  In particular, and generally in
order of increasing complexity, we will deal with abstracting the
domain, the matrix representation, the code itself, determinant, etc.
Then we put it all together (or in a new section?).

@@

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
\jacques{I don't think I really want to delete the first half of this
subsection.  I guess we can move that into section 3?  We need to be
careful that section 3 does not become 1/2 the paper!  We might even
consider splitting section 3 into 2 sections?}
We will be using MetaOCaml operations to build code expressions, values
of type |('a,'v) code|. However, we wish to make our generators
abstract over particular code representations. Therefore, when writing
the LU generator (Section~\ref{sec:det} and below), instead of
MetaOCaml brackets and escapes we will be using the code combinators
described below. These combinators are functions that produce and
combine values of the type of \emph{abstract code}.

@@

\subsection{Determinant aspect}
\label{sec:det}
% Various aspects of our LU generator such as determinant, rank, 
% and permutation tracking may need to keep a state during
% code generation. For example, the initializing section of the LU
% generator will invoke a function (|decl|) of the determinant aspect 
% which ``declares'' whether tracking the determinant is needed, 
% so that the
% latter may generate let-binding for reference cells that track the
% magnitude and the sign of the determinant. The generator of the
% row-swapping code will notify the determinant aspect so that the
% latter may insert code that flips the sign of the tracked
% determinant. The final section of the LU generator will ask the
% determinant aspect to generate code producing the final, signed value
% of the determinant. Clearly the determinant aspect needs to maintain
% state from one invocation of its functions to another; this state should
% include the ``names'' of mutable variables accumulating the sign and the
% magnitude of the determinant.

To track the determinant, we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign (|upd_sign|) or the absolute
value of the determinant, and converting the tracking state
to the final determinant value of type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| (see definition below) where |outdet| is |unit|
and all the
functions of that module generate no code. For integer matrices, and
in general whenever the matrix elements do not form a field, we
have to track some aspects of the determinant, even if we do not output
it, as the fraction-free update needs those aspects. The determinant tracking
aspect is complex because tracking variables, if any, are to be declared at the
beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. 
Other aspects follow a similar outline.
